Stochastic Approximation Monte Carlo with Dynamic Update Factor (SAD)

### 2018-08-03 papers/histogram/sad.tex
reworked wltmmc section in the paper.

\begin{document}
\title{Dynamic Stochastic Approximation Monte Carlo}

@@

\davidsays{We should discuss continuous energies and binning somewhere
  here...}
\jpsays{I put some text about this but it was under a sad subsection also
may need reworked and/or edited.}

@@

\subsection{TMMC Algorithm}
The Transition Matrix Monte-Carlo (TMMC) method as introduced by
Fitzgerald et al. consists of three primary steps.  The first two steps
are common to the Metropolis algorithm while the third makes use of the
actual transition probabilities~\cite{fitzgerald2000monte}.

\subsubsection{define a transition probability}
Assuming that the system under study is in a state denoted old, and a
proposed transition is made to a new state denoted new, the probability
is defined as
\begin{align}
  p_{old \rightarrow new} = p_{new \rightarrow old}
\end{align}
where we have allowed the probabilies to be equal for simplicity of
presentation.

\subsubsection{define an acceptance probability}
The probability for accepting a transition from an old state to a new
one can be defined
\begin{align}
  p_{a} = \min\bigg[1,\frac{\alpha_{new\rightarrow old}}
  {\alpha_{old \rightarrow new}}\frac{\pi_{new}}{\pi_{old}}\bigg]
\end{align}
where $\alpha_{new\rightarrow old}$ is the probability of generating a
new system configuration the old.  For conventional Monte-Carlo moves
$\alpha_{old \rightarrow new} =\alpha_{new\rightarrow old}$ although
for advanced Monte-Carlo moves this would not be the
case~\cite{paluch2008comparing, siepmann1990method}.  The probabily of
observing the system in the old state can be found based on the
ensemble constraints placed on the system.  Fitzgerald et al.
considered the system of interest using the canonical ensemble;
however, the extension to other ensembles is quite natural.
\begin{align}
  \pi_{old} = \frac{e^{-\beta E_{old}}}{Z}
\end{align}
\subsubsection{define a bookkeeping step}
The third step in the TMMC algorithm consists of recording the data in
a collection matrix (the unormalized transition histogram matrix).
\begin{align}
  C_{N,\delta} = C_{N,\delta} + p_{a}\\
  C_{N,0} = C_{N,0} +(1 - p_{a})
\end{align} If $\delta=0$, that is, the energy state has not been
changed, then only $C_{N,0}$ is incremented by unity. Fitzgerald et
al.~\cite{fitzgerald2000monte} showed that defining the third step this
way results in a uniform improvement over using histogram estimators to
determine the particle number probability distribution (PNPD).

\subsection{WL-TMMC Algorithm}

Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC) method as proposed
by Shell and coworkers~\cite{shell2003improved,shell2004flat} takes
advantage of the ease of implementation and applicability to a variety
of systems, while making use of a transition matrix to address WL
weaknesses, such as its limited statistical accuracy plateau.
This WL version is more accurate for a given number of simulation steps.

\paragraph{Initializing Wang-Landau:} We begin the simulation by
defining a minimum important energy and a maximum entropy state. In
traditional Wang-Landau the WL factor begins at 1.0 and the factor is
modified by 2.0 until the simulation is cutoff typically around
$10^{-10}$.  Rane et. al. published an implemenation of
WL-TMMC~\cite{rane2013monte} that suggested using a WL factor between
1.0 and $10^{-2}$ with a cutoff $<10^{-5}$. In our implementation of
WL-TMMC, we set the WL factor to be 1 and the cutoff to be either
$10^{-4}$ or $10^{-10}$. A single sweep is defined as each macrostate
being sampled a number of times in this case 10,000 (although 100 is
often sufficient for small systems). Wang-Landau sets a flatness
criteria~\cite{wang2001determining, wang2001efficient,
hatch2015computational, mahynski2017predicting} for the accumulated
histogram usually between 0.75 to 0.90.  In the original
implementation, it was found to be sufficient that each macrostate is
visited at least once (each histogram bin has one
entry)~\cite{shell2003improved}; however, for other systems this is
rarely sufficient.  The goal of this WL initialization is to prefill
the transition matrix for the next portion of the simulation.
\paragraph{Initializing TMMC:} Prefilling the transition
matrix is necessary because if numerous zeros existed in the collection
matrix, the infinite temperature transition matrix would be
ill-defined.  Sampling over large ranges of density of states would
therefore take an unacceptable length of time to
complete~\cite{shell2003improved, shen2014elucidating}.  The transition
matrix in terms of the collection matrix is given as
follows:
\begin{align}
\widetilde{T}_{\infty}(I\rightarrow J) = \frac{C(I,J)}
{\sum_{K} C(I,K)}
\end{align}
We now run the TMMC portion of the simulation.
%Siderius does this for 20 additional sweeps (does about 8 sweeps for
%WL).  We do this until we reach a defined number of iterations. We
%have not yet implemented a biasing function to control acceptance rate
%of trial moves but we could? at some point.
The acceptance probability is written in terms of the infinite temperature
transition matrix.
\begin{align}
  P_{acc}(i\rightarrow j) = \min\bigg[1,\frac{\widetilde{T}_{\infty}(J\rightarrow I)}
  {\widetilde{T}_{\infty}(I\rightarrow J)}\bigg]
\end{align}
This probability is only used to control actual transitions not update
the collection matrices.

\section{SAD Algorithm}\label{sec:sad}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) method is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Flat-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the user to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

@@

\subsubsection{The ``interesting'' energy range}\label{sec:EHL}
Since SAD explores all energy states it needs to have a reasonable idea
of what energy range corresponds to the temperature range of interest,
defined by $T_{\min}<T<\infty$.
The
simulation is responsible for determining and updating this energy
range.
Given the true entropy $S(E)$, we can define the interesting energy
range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  In the course of the
simulation this precise energy may be challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$ to define the range over which the energy histogram
is made flat. $E_H$ is chosen to be the greatest energy that
maximized the entropy at any time during the simulation.  Similarly,
$E_L$ is the lowest energy that ever maximized $S(E)-E/T_{\min}$ during
the simulation.  This creates a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

@@

\subsubsection{Energy binning with SAD}

@@

\subsubsection{Scaling, step size, and SAD}
Methods such as WL can fail to converge for complex systems in part
because the same random MC moves are performed for the entire energy
range.  The main problem lies in trying to determine how to optimize
the step size such that the time to explore all energies is reasonable.
Small step sizes lead to improved sampling at the expense of slow
convergence. Koh et al. proposed multiplying $\gamma$ by an energy
dependent factor in order to improve the converge of WL in regards to
convergence and scaling~\cite{koh2013dynamically}.  Because the update
factor for SAD has an energy dependent factor already present (and
dynamically adjusted), we expect that SAD will perform consitently
scaling free for a reasonable choice of step size.  SAD should perform
extremely well compared with other flat histogram methods when applied
to complex continuous systems such as polymers/proteins and spin models.

>>>>

\begin{document}
\title{Stochastic Approximation Monte Carlo with Dynamic Update
Factor (SAD)
}

@@

{\color{red} Both weight-based and transition matrix methods calculate
the density of states $D(E)$ for discrete energy
levels~\cite{haber2018performance}. For this reason, energy binning
becomes an important consideration for complex continous systems.
Energy bins are typically of uniform size for the entire energy
continuum~\cite{fasnacht2004adaptive}. Coarse binning yields fewer
energy states to explore but more time spent in each individual bin.
Fine binning yields more energy states to explore with less time spent
in each bin. Some methods such as AdaWL~\cite{koh2013dynamically}
employ a tunable mechanism for controlling the binning for low entropic
states in order to ensure the exploration of all energies. }

@@

\subsection{TMMC Algorithm}
The previously described weight-based flat-histogram methods utilizes a so-called
``visited states'' approach to calculate  thermodynamic
properties~\cite{errington2003direct}. The algorithm is fundamentally
based on the number of times a given state is visited while the
system is being sampled. The Transition Matrix Monte Carlo (TMMC) method operates
differently than the visited states approach~\cite{fitzgerald2000monte}.
The method uses attempted transitions between states to obtain an estimate of
the probability that a given transition will be proposed.  The density
of states is computed from the resulting transition matrix.  An advantage
of collecting the transition matrix is that this matrix may be collected
while another method is employed to compute acceptance probabilities, as
in Wang-Landau TMMC, discussed below.

The TMMC algorithm records the transition data in a \emph{collection matrix},
i.e. the unnormalized transition histogram matrix, which stores the number
of times a transition between two energy states has been proposed.
\begin{align}
  C(E_{\text{old}} \rightarrow E_{\text{new}},t+1)
     = C(E_{\text{old}} \rightarrow E_{\text{new}},t) + 1
\end{align}
The transition matrix is computed from the collection matrix via
\begin{align}
  T(E_{\text{old}} \rightarrow E_{\text{new}})
   = \frac{C(E_{\text{old}} \rightarrow E_{\text{new}})}{
     \sum_E C(E_{\text{old}} \rightarrow E)
  }
\end{align}
The acceptance probability is defined in terms of this biasing function:
\begin{align}
  \mathcal{P}(E_{\text{old}} \rightarrow E_{\text{new}}) =
  \max\left(1,\frac{T(E_{\text{old}} \rightarrow E_{\text{new}})}{
       T(E_{\text{new}} \rightarrow E_{\text{old}})}\right)
\end{align}
where we note that the transition matrix is updated to reflect the
transition under consideration \emph{prior} to computing the acceptance
probability.  The TMMC method suffers from poor convergence primarily
because it is sensitive to poor statistics in less probable energy
transitions.  Like the weight-based methods above, the TMMC method violates
detailed balance, with the level of violation decreasing as the simulation
progresses and a difference of 1 in the collection matrix has a smaller
impact on the acceptance probabilities.

\subsection{WL-TMMC Algorithm}
Wang-Landau Transition Matrix Monte-Carlo (WL-TMMC) method as proposed
by Shell and coworkers~\cite{shell2003improved,shell2004flat} takes
advantage of the WL methods ease of implementation, while making use of
a transition matrix to address WL weaknesses, such as its limited
statistical accuracy convergence.

\paragraph{Initializing Wang-Landau:}
We set the energy range along which the method may sample as
$E(T_{\min}) <E< E(T=\infty)$. Conventional Wang-Landau prescribes
updating the WL factor $\gamma$ by starting at 1.0 and reducing by
$\frac12$ until the simulation is cutoff typically around $10^{-10}$.
While variants of this approach appear in the literature, Rane et. al.
published an implemenation of WL-TMMC~\cite{rane2013monte} that
suggested using a WL factor between 1.0 and $10^{-2}$ with a cutoff
$<10^{-5}$, We implement WL-TMMC by setting the WL factor to be 1.0 and
the cutoff to be either $10^{-4}$ or $10^{-10}$.

A single sweep ``roundtrip'' is defined as each macrostate being
sampled a number of times in this case 100,000 times. Wang-Landau sets
a flatness criteria~\cite{wang2001determining, wang2001efficient,
hatch2015computational, mahynski2017predicting} for the accumulated
histogram usually to be 0.8.  In the original implementation, it was
found to be sufficient that each macrostate is visited at least once
(each histogram bin has one entry)~\cite{shell2003improved}; however,
for other systems this is rarely sufficient~\cite{bhar2009computer}.
The goal of this WL initialization is to prefill the transition matrix
for the next portion of the simulation.

\paragraph{Initializing TMMC:} Prefilling the transition matrix is
necessary because if numerous zeros existed in the collection matrix,
the infinite temperature transition matrix would be ill-defined.
Sampling over large ranges of density of states would therefore take an
unacceptable length of time to complete~\cite{shell2003improved,
shen2014elucidating}.  The TMMC portion of the algorithm is run only
after WL has reached the specified cutoff. An important note is that
the acceptatnce probability for TMMC is only used to control actual
transitions not update the collection matrices.

\section{SAD Algorithm}\label{sec:sad}
The Stochastic Approximation with Dynamic update factor (SAD) method
is a variant of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Flat-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the user to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

@@

Since SAD does not explore \emph{all} energy states it needs to have a
reasonable idea
of what energy range corresponds to the temperature range of interest,
defined by $T_{\min}<T<\infty$.
The
simulation is responsible for determining and updating this energy
range.
Given the true entropy $S(E)$, we can define the interesting energy
range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  In the course of the
simulation this precise energy may be challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$ to define the range over which the energy histogram
is made flat. $E_H$ is chosen to be the greatest energy that
maximized the entropy at any time during the simulation.  Similarly,
$E_L$ is the lowest energy that ever maximized $S(E)-E/T_{\min}$ during
the simulation.  This creates a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

@@

\paragraph{Energy binning with SAD}

@@

\paragraph{Scaling, step size, and SAD}
Methods such as WL can fail to converge for complex systems in part
because the same random MC moves are performed for the entire energy
range.  The main problem lies in trying to determine how to optimize
the step size such that the time to explore all energies is reasonable.
Small step sizes lead to improved sampling at the expense of slow
convergence. Koh et al. proposed multiplying $\gamma$ by an energy
dependent factor in order to improve the converge of WL in regards to
convergence and scaling~\cite{koh2013dynamically}.  Because the update
factor for SAD has an energy dependent factor already present (and
dynamically adjusted), we expect that SAD will perform consitently
scaling free for a reasonable choice of step size.  SAD should perform
extremely well compared with other flat histogram methods when applied
to complex continuous systems such as polymers/proteins and spin models.
